The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-08 thesis/extra_density.tex
grammar/spelling edits

\chapter{The reliability of derived density
profiles}\label{sec:extra_density}

In this section, we detail the J+24 membership algorithm and test the
resulting density profiles for Scl and UMi. We compare profiles derived
with alternative methodologies and literature results. We also present a
non-parametric Bayesian density profile, which reproduces J+24's results
in the inner regions but diverges when the background dominates. In
summary, we find the density profiles are robust up to the
background-limited radius derived here.

@@

To create a high-quality sample, J+24 select stars initially from Gaia
within a 2--4 degree region around each satellite with high-quality
astrometry, reliable photometry, and consistent parallaxes and broadly
consistent proper motions and colours. Stars are removed if they have
excess astrometric noise \citep[\({\rm ruwe} > 1.3\),
see][]{lindegren+2021}, colour excess \(3\sigma\) outside of
expectations \citep[from][]{riello+2021}, proper motions magnitudes
\(>10\,\masyr\) in \(\mu_{\alpha*}\) or \(\mu_\delta\), magnitudes
brighter than the tip of the red giant branch or fainter than \(G=22\),
or colours outside \(-0.5 < G_{\rm BP} - G_{\rm RP} <  2.5\). Photometry
is dereddened with \citet{schlegel+finkbeiner+davis1998} extinction
maps.

@@

The PM likelihood is a bivariate gaussian with variance and covariance
equal to each star's proper motions.

J+24 model the CMD as a Padova isochrone \citep{girardi+2002}, with
matching metallicity, 12 Gyr age (or 2 Gyr for Fornax), and Gaussian
width of 0.1 mag plus the \emph{Gaia} colour uncertainty at each
magnitude. The horizontal branch is modelled as a
constant-magnitude\footnote{Specifically, the mean magnitude of the 12
  Gyr, \({\rm [Fe/H]}=-2.2\), Padova isochrone's horizontal branch.}
sequence extending blue of the CMD with the same width as the RGB. The
CMD likelihoods are marginalized over the distance modulus and take the
maximum of the RGB and horizontal branch likelihoods.

The background likelihoods are determined empirically as a kernel
density estimates from stars outside 5\(R_h\) in PM and CMD space. The
spatial background likelihood is uniform.

@@

@@

In this section, we test how assumptions in J+24's algorithm or biases
in \emph{Gaia} may affect density profiles. In all cases, density
profiles converge until \(R_{\rm ell} \approx 60\,{\rm arcmin}\), where
background contamination likely dominates.

\textbf{Spatial likelihood.} As J+24 include a spatial likelihood, our
density profiles risks ``double fitting''---the spatial distribution of
stars informs the membership catalogue from which density profiles are
derived. If the assumed spatial likelihood is misrepresentative, the
derived density profiles may bias towards the assumed likelihood.
Fig.~\ref{fig:scl_umi_density_extras} illustrates that selecting the
one- or two-component spatial models affects the outer extent of the
dwarf galaxy. Promisingly, the density excess persists when assuming a
one-component exponential profile. We revisit this concern using
spatially independent methods below.

\textbf{Structural uncertainties}. The assumed structural parameters of
a dwarf (centre, position angle, ellipticity, scale radius) may be
biased or vary radially. As a test, we use a membership sample J+24
derive assuming the outer component is circular. Even when binning in
circular radius, the density profiles remain unchanged (the
\texttt{circ} points in Fig.~\ref{fig:scl_umi_density_extras}). Thus,
our conclusions appear to be robust to an extreme shift in ellipticity.
Reasonable changes in other structural parameters are unlikely to have a
stronger effects.

\textbf{Completeness}. \emph{Gaia} appears to have high but imperfect
completeness, more limited for crowded fields, faint sources
(\(G\gtrsim20\)), and BP-RP magnitudes. In \citet{fabricius+2021}, the
completeness down to \(G\approx 20\) is \(\sim 80\%\) for low-density
globular clusters. As dwarfs are even less dense, their completeness is
likely higher. To test for magnitude-dependent biases, we show density
profiles derived for only the brighter half of member stars in
Fig.~\ref{fig:scl_umi_density_extras}. We find no substantive change
compared to our fiducial density profile. If \emph{Gaia}'s
incompleteness is inhomogeneous, then these variations likely do not
affect our density profiles.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/density_methods_extra.pdf}}
\caption[Sculptor density methodology comparison]{Density profiles for
various assumptions for Sculptor. \texttt{2-exp} is the fiducial
double-exponential-likelihood J+24 sample, \texttt{1-exp} instead is a
one-component exponential spatial likelihood, \texttt{simple} uses the
simple position-independent selection criteria, \texttt{circ} is a
2-component bayesian model assuming circular radii, \texttt{bright} only
includes stars brighter than the median magnitude, and \texttt{DELVE} or
\texttt{UNIONS} use photometry from external surveys with background
subtraction. In all cases, the density profiles are nearly identical
until the background-limited regime (grey shaded
region).}\label{fig:scl_umi_density_extras}
\end{figure}

\section{Comparison against alternate
samples}\label{sec:simple_selection}

Next, to compare against simpler methods, we create samples using
absolute filters in the CMD, PM space. We also compare against similarly
selected samples from deep photometry-only data.

For the \texttt{simple} \emph{Gaia} samples, we require high-quality
astrometry (\texttt{ruwe} \textless{} 1.3), parallaxes
\(3\sigma\)-consistent with the dwarf's distance, proper motions with
\(1\masyr\) radius of the galaxy's value, and within the dwarf's
empirical CMD (Fig.~\ref{fig:extra_cmd}).

We also determine density profiles from the deeper photometric surveys.
For Scl, we use DELVE DR2 survey \citep{drlica-wagner+2022}. We select
sources within an ellipse of radius 150 arcminutes, categorized as
likely stars, with reliable \(g\) and \(r\) magnitudes (associated flags
\(\leq 4\)), and within Scl's observed CMD (Fig.~\ref{fig:extra_cmd}).
Ursa Minor is within the Ultraviolet Near-Infrared Optical Northern
Survey \citep[UNIONS,][]{gwyn+2025}. We select sources within an ellipse
of radius 230 arminutes, with no \texttt{FLAGS\_CFIS} set, the
\texttt{s21} and \texttt{s31} both \(<3\) (i.e.~not extended sources),
and within the CMD cut in Fig.~\ref{fig:extra_cmd}.
Fig.~\ref{fig:delve_unions_tangent} shows the resulting spatial
distribution for both samples. The only remarkable structure is Muñoz 1
nearby to Ursa Minor.

Finally, Fig.~\ref{fig:scl_umi_density_extras} compares the density
profiles from each sample. All samples agree until the ``limiting
radius'', where the \texttt{simple} and DELVE/UNIONS samples reach their
respective backgrounds. The very outskirts of J+24's density profiles
may also extend below the background of apparent member stars,
complicating the reliability of the outer density profiles.

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/extra_cmd_selection.png}
\caption[Colour-Magnitude sample selection]{Colour-magnitude diagram
cuts used in samples in this section. The likely-member stars in the
inner \(1R_h\) of each dwarf (black points) are used as a guide to
determine the CMD cuts (orange polygons). Light grey points instead show
the full distribution of stars across the field.}\label{fig:extra_cmd}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/delve_unions_tangent.pdf}}
\caption[DELVE and UNIONS spatial distribution of stars]{The
distribution of DELVE and UNIONS selected stars in Scl and UMi as grey
points overdrawn with isophotes. The clump to the lower right of UMi is
Muñoz 1.}\label{fig:delve_unions_tangent}
\end{figure}

@@

To address the concerns discussed above, we consider here a
non-parametric model to fit the density in each bin. We demonstrate that
the non-parametric model fails to find evidence for stellar density
where J+24 still detects members. Notably, our density profiles diverge
substantially from J+24 for Antlia II. We suggest these differences
arise from background contamination, motivating our ``limiting radii''
representing where densities may become unreliable.

@@

As a non-parametric but similar model to J+24, we consider a piecewise
constant spatial likelihood. In this model, the stars are divided into
radial bins which are then considered independently. Each bin has only
one free parameter, the fraction of satellite stars in that bin,
\(f_{\rm sat}\). This model thus directly derives the density profile
from the data in a single step. If a bin contains insufficient
information to estimate a precise satellite density, than the posterior
\(f_{\rm sat}\) should reflect this uncertainty.

We parameterize \(f_{\rm sat}\) in terms of a log-relative density,
\(\theta\) \begin{equation}{
\begin{split}
\theta &\equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}}) \\
f_{\rm sat} &= \frac{10^{\theta}}{1 + 10^{\theta}}.
\end{split}
}\end{equation} We then adopt a broad uniform prior on \(\theta\) from
-12 to 6. The CMD and PM likelihoods are unchanged from J+24, except the
systemic PM is fixed for efficiency.

@@

\begin{table*}[t]
\centering
\caption[The limiting radii of Gaia-derived density profiles]{For each classical dwarf, the limiting radius $R_{\rm limit}$ in units of $R_h$ and arcminutes. $R_{\rm limit}$ represents where there no longer appears to be evidence of stars in Emph(Str(Gaia)) using the nonparametric MCMC density profiles. }
\label{tbl:mcmc_props}
\begin{tabular}{lll}
\toprule
Galaxy & $R_{\rm limit} / R_h$ & $R_{\rm limit} / '$\\
\midrule
Fornax & 5.25 & 79.1\\
Sculptor & 6.39 & 64.1\\
Leo I & 4.23 & 13.5\\
Ursa Minor & 6.42 & 86.4\\
Leo II & 3.63 & 8.76\\
Carina & 4.16 & 33.3\\
Draco & 3.59 & 27.3\\
Canes Venatici I & 1.95 & 12.5\\
Sextans I & 3.42 & 67.9\\
Crater II & 1.93 & 39.0\\
\bottomrule
\end{tabular}
\end{table*}

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived, MCMC
density profiles as compared to J+24. In general, both methodologies are
consistent. However, J+24 tend to systematically overestimate faint
densities and confidently derive densities where the MCMC model fails to
derive a density estimate.

In the outskirts of satellites, more background / foreground stars may
have consistent PM and CMD properties than satellite members. Improperly
estimating the satellite's density in this ``background-limited'' regime
likely affects the inferred density of members. If the satellite's
density is severely overestimated, then a J+24-like sample may select
many additional background stars. The density of candidates would then
be biased towards the assumed density. As J+24 only use a one or
two-component density profile across the entire dwarf, the density
profile is likely biased where the CMD+PM background dominates. The
nonparametric MCMC model instead does not assume a local density, so
should better represent the underlying satellite density.

@@

Finally, we compare our density profiles against a sample of
literature-derived density profiles in
Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. Deviations
between different profiles are slight, despite the range of methods
across time, and all density profiles extending into the outskirts of
Scl and UMi show a similar overdensity to the ones we find. The extended
stellar profiles of Scl and UMi appear to be a robust result across the
literature.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Sculptor literature density profiles]{A comparison of the Scl
derived density profile and historical works. Unlike most profiles in
this thesis, this density profile is plotted with respect to the
semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
The solid black line is a 2D exponential with corresponding scale radius
to Scl's half-light radius, and the residuals in the bottom panel are
with respect to this profile. References are (in order),
\citet{munoz+2018}; \citet{westfall+2006}; \citet{walcher+2003};
\citet{eskridge1988}; \citet{demers+krautter+kunkel1980}; and
\citet{hodge1961}.}\label{fig:scl_lit_profiles}
\end{figure}

@@

In this Appendix, we discussed the methodological details of the J+24
sample selection algorithm. We consider possible biases due to the
assumed spatial likelihood, \emph{Gaia}'s completeness, and structural
parameters. In all cases, we find these assumptions likely do not cause
major biases in the derived density profiles. However, we show that
J+24's density profiles may become unreliable when dropping below the
background of satellite-like stars. Finally, comparing our density
profiles against the literature, we find our results to be consistent.
We conclude that the detection of an extended density profile in Scl and
UMi is robust to incompleteness, methodology, alternative surveys, and
across the literature.

>>>>

\chapter{\texorpdfstring{The Reliability of Projected Number Density
Profiles Derived from
\emph{Gaia}}{The Reliability of Projected Number Density Profiles Derived from Gaia}}\label{sec:extra_density}

In this section, we detail the J+24 membership algorithm and test the
resulting density profiles for Scl and UMi. We compare profiles derived
with alternative methodologies and literature results. We also present a
non-parametric Bayesian density profile, which reproduces J+24's results
in the inner regions but differs when the background dominates. We find
the density profiles are robust up to the background-limited radius
derived here.

@@

To create a high-quality sample, J+24 select stars initially from Gaia
within a 2--4 degree region around each satellite with high-quality
astrometry, reliable photometry, consistent parallaxes, and broadly
consistent proper motions and colours. Stars are removed if they have
excess astrometric noise \citep[\({\rm ruwe} > 1.3\),
see][]{lindegren+2021}, colour excess \(3\sigma\) outside of
expectations \citep[from][]{riello+2021}, proper motions \(>10\,\masyr\)
in \(|\mu_{\alpha*}|\) or \(|\mu_\delta|\), magnitudes brighter than the
tip of the red giant branch\footnote{Derived from the maximum RGB
  magnitude from the associated Padova isochrone plus
  \(5\sigma\)\hspace{0pt} times the distance modulus uncertainty.} or
fainter than \(G=22\), or colours outside
\(-0.5 < G_{\rm BP} - G_{\rm RP} <  2.5\). Photometry is dereddened with
\citet{schlegel+finkbeiner+davis1998} extinction maps.

@@

The PM likelihood is a bivariate Gaussian with variance and covariance
equal to each star's proper motions.

J+24 model the CMD as a Padova isochrone \citep{girardi+2002}, with
metallicity matching that obtained spectroscopically, 12 Gyr age (or 2
Gyr for Fornax), and a Gaussian width of 0.1 mag plus the \emph{Gaia}
colour uncertainty at each magnitude. The horizontal branch is modelled
as a constant-magnitude\footnote{Specifically, the mean magnitude of the
  12 Gyr, \({\rm [Fe/H]}=-2.2\), Padova isochrone's horizontal branch.}
sequence extending blue of the CMD with the same width as the RGB. The
CMD likelihoods are marginalized over the distance modulus and take the
maximum of the RGB and horizontal branch likelihoods.

The background likelihoods are determined empirically from kernel
density estimates of stars outside 5\(R_h\) in PM and CMD space. The
spatial background likelihood is uniform.

@@

\section{Independent sample
selection}\label{independent-sample-selection}

To test for possible limitations in the J+24 methodology (discussed
below), we compare their density profiles against samples derived from
absolute cuts in \emph{Gaia} and external photometric surveys.

The \texttt{simple} sample is selected from stars in \emph{Gaia} using
absolute astrometric and quality cuts. We require high-quality
astrometry (\texttt{ruwe} \textless{} 1.3), parallaxes
\(3\sigma\)-consistent with the dwarf's distance, proper motions with
\(1\masyr\) of the systemic value, and colours within the dwarf's
apparent CMD as determined from stars near the dwarf's centre (see
Fig.~\ref{fig:extra_cmd}).

We also determine density profiles from the deeper photometric surveys.
For Scl, we use data from the DECam Local Volume Exploration Survey
(DELVE) DR2 survey \citep{drlica-wagner+2022}. We select sources within
an ellipse of radius 150 arcminutes, categorized as likely stars, with
reliable \(g\) and \(r\) magnitudes (associated flags \(\leq 4\)), and
within Scl's CMD selection (orange polygon in Fig.~\ref{fig:extra_cmd}).
Ursa Minor lies within the Ultraviolet Near-Infrared Optical Northern
Survey \citep[UNIONS,][]{gwyn+2025}. We select sources within an ellipse
of radius 230 armin, with no \texttt{FLAGS\_CFIS} set, both \texttt{s21}
and \texttt{s31} \(<3\) (i.e., not extended sources), and within the CMD
polygon in Fig.~\ref{fig:extra_cmd}.

\begin{figure}
\centering
\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{figures/extra_cmd_selection.png}
\caption[Colour-Magnitude sample selection]{Colour-magnitude diagram
cuts used in samples in this section. Black points represent stars
within \(1R_h\), light grey shows all field stars passing quality cuts,
and the orange polygon shows the CMD cut. The central, likely-member
stars inform the CMD selections.}\label{fig:extra_cmd}
\end{figure}

@@

We now discuss potential biases in J+24's algorithm for \emph{Gaia}
data. We then test how these biases may influence density profiles. In
all cases, density profiles converge within
\(R_{\rm ell} \approx 60\,{\rm arcmin}\), where background contamination
begins to dominate.

\textbf{Spatial likelihood.} Since J+24 include a spatial likelihood,
the resulting density profiles risk ``double fitting''---the same
spatial distribution of stars informs the membership catalogue and the
subsequent density profiles. If the spatial likelihood misrepresents the
underlying satellite structure, then the density profiles may be biased
towards the assumed shape.

\textbf{Structural uncertainties.} The assumed structural parameters of
a dwarf (centre, position angle, ellipticity, scale radius) may be
inaccurate or vary radially. This would change the elliptical radii of
stars (\(R_{\rm ell}\)) and likewise the density profiles.

\textbf{Completeness.} \emph{Gaia}'s completeness appears high but
imperfect, more limited in crowded fields, for faint sources
(\(G\gtrsim20\)), and in \(G_{\rm BP}\)-\(G_{\rm RP}\) magnitudes. In
\citet{fabricius+2021}, the completeness down to \(G\approx 20\) is
\(\sim 80\%\) for low-density globular clusters. As dwarfs are even less
dense, their completeness is likely higher.

The top row of Fig.~\ref{fig:scl_umi_density_extras} compares the
density profiles assuming different spatial likelihoods: the
\texttt{2-exp} and \texttt{1-exp} use two or one-component spatial
likelihoods in J+24, and \texttt{simple} does not assume a spatial
likelihood (described above). All profiles agree within the
background-limited (BG-limited) radius, derived below. The
\texttt{1-exp} profile predicts lower outer densities, yet the outer
density excess in Sculptor and Ursa Minor persists. Thus, the spatial
likelihood can influence the outer regions but has less effect on the
inner regions.

The lower panels of Fig.~\ref{fig:scl_umi_density_extras} test other
possible biases. The \texttt{circ} profile uses circular bins and the
J+24 sample, assuming a circular outer profile. As the profile is nearly
identical to the fiducial model, an extreme change in ellipticity does
not affect our conclusions. The \texttt{bright} profile is derived using
only the brighter half of stars. We find no substantive change compared
to our fiducial density profiles, showing completeness-related magnitude
biases are likely small. Finally, the DELVE and UNIONS density profiles
nearly identically trace the \texttt{2-exp} profile, but both become
background-limited near the BG-limited radius.

In all cases, there is little evidence for biases in \emph{Gaia}'s
completeness or sensitivity to structural parameters. As the spatial
likelihood appears to be the most severe bias, we next consider a more
flexible model.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/density_methods_extra.pdf}}
\caption[Density methodology comparison]{Density profiles for various
assumptions for Sculptor (left) and Ursa Minor (right). Each profile is
plotted as points in log surface density (scaled to the \texttt{2-exp}
profile) versus log radius, the black line marks a 2D exponential
profile, and the grey shaded region represents the background-limited
region derived in
Section~\ref{sec:mcmc_hists}.}\label{fig:scl_umi_density_extras}
\end{figure}

@@

To address the concerns discussed above, we consider here a
non-parametric model to fit the density in each bin. We demonstrate that
the non-parametric model fails to find evidence for stellar density,
where J+24 still detects members. Notably, our density profiles diverge
substantially from J+24 for Antlia II. We suggest these differences
arise from background contamination, motivating our ``limiting radii''
representing where densities may become unreliable.

@@

As a non-parametric but similar model to J+24, we consider a piecewise
constant spatial likelihood. In this model, the stars are divided into
radial bins, which are then considered independently. Each bin has only
one free parameter, the fraction of satellite stars in that bin,
\(f_{\rm sat}\). This model thus directly derives the density profile
from the data in a single step. If a bin contains insufficient
information to estimate a precise satellite density, then the posterior
\(f_{\rm sat}\) should reflect this uncertainty.

We parameterize \(f_{\rm sat}\) in terms of a log-relative density,
\(\theta\) \begin{equation}{
\begin{split}
\theta &\equiv \log_{10}({\Sigma_{\rm sat}}/{\Sigma_{\rm bg}}) \\
f_{\rm sat} &= \frac{10^{\theta}}{1 + 10^{\theta}}.
\end{split}
}\end{equation} We then adopt a broad uniform prior on \(\theta\) from
-12 to 6. The CMD and PM likelihoods are unchanged from J+24, except
that the systemic PM is fixed for efficiency.

@@

\begin{table*}[t]
\centering
\caption[The limiting radii of Gaia-derived density profiles]{For each classical dwarf, the limiting radius $R_{\rm limit}$ in units of $R_h$ and arcminutes. $R_{\rm limit}$ represents where there is no longer evidence of Emph(Str(Gaia)) members using the nonparametric MCMC density profiles. }
\label{tbl:mcmc_props}
\begin{tabular}{lll}
\toprule
Galaxy & $R_{\rm limit} / R_h$ & $R_{\rm limit} / '$\\
\midrule
Fornax & 5.25 & 79.1\\
Sculptor & 6.39 & 64.1\\
Leo I & 4.23 & 13.5\\
Ursa Minor & 6.42 & 86.4\\
Leo II & 3.63 & 8.76\\
Carina & 4.16 & 33.3\\
Draco & 3.59 & 27.3\\
Canes Venatici I & 1.95 & 12.5\\
Sextans I & 3.42 & 67.9\\
Crater II & 1.93 & 39.0\\
\bottomrule
\end{tabular}
\end{table*}

Figs.~\ref{fig:mcmc_hists}, \ref{fig:mcmc_hists2} show the derived MCMC
density profiles as compared to J+24. In general, both methodologies
produce similar densities. However, J+24 tend to systematically
overestimate faint densities and confidently derive densities where the
MCMC model fails to derive a density estimate.

In the outskirts of satellites, more background/foreground stars may
have consistent PM and CMD properties than satellite members. Improperly
estimating the satellite's density in this ``background-limited'' regime
likely affects the inferred density of members. If the satellite's
density is severely overestimated, then a J+24-like sample may select
many additional background stars. The density of candidates would then
be biased towards the assumed density. As J+24 only use a one or
two-component density profile across the entire dwarf, the density
profile is likely biased where the CMD+PM background dominates. The
nonparametric MCMC model instead does not assume a local density, so it
should better represent the underlying satellite density.

@@

Finally, we compare our density profiles against a sample of
literature-derived density profiles in
Figs.~\ref{fig:scl_lit_profiles}, \ref{fig:umi_lit_profiles}. Deviations
between different profiles are small, despite the range of methods
across time, and all density profiles extending into the outskirts of
Scl and UMi show a similar overdensity to the ones we find. The extended
stellar profiles of Scl and UMi appear to be a robust result across the
literature.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_literatre_profiles.pdf}}
\caption[Sculptor literature density profiles]{A comparison of density
profiles of Sculptor for works from the literature. Unlike most profiles
in this thesis, this density profile is plotted with respect to the
semi-major elliptical radius (\(a = R_{\rm ell} / \sqrt{1-{\rm ell}}\)).
The solid black line is a 2D exponential with a corresponding scale
radius to Scl's half-light radius, and the residuals in the bottom panel
are with respect to this profile. References are (in order),
\citet{munoz+2018}; \citet{westfall+2006}; \citet{walcher+2003};
\citet{eskridge1988}; \citet{demers+krautter+kunkel1980}; and
\citet{hodge1961}.}\label{fig:scl_lit_profiles}
\end{figure}

@@

In this Appendix, we discussed the methodological details of the J+24
sample selection algorithm. We consider possible biases due to the
assumed spatial likelihood, \emph{Gaia}'s completeness, and structural
parameters. In all cases, we find that these assumptions likely do not
cause major biases in the derived density profiles. However, we show
that J+24's density profiles may become unreliable when dropping below
the background of satellite-like stars. Finally, comparing our density
profiles against the literature, we find our results to be consistent.
We conclude that the detection of an extended density profile in Scl and
UMi is robust to incompleteness, methodology, alternative surveys, and
across the literature.

### 2025-10-08 thesis/extra_numerical_convergence.tex
grammar/spelling edits

where \(h_{\rm grav}\) is the softening length, and \(N_{200}\) is the
number of particles within \(R_{200}\). This choice balances integration
time and only compromises resolution in collisional regime. For our
isolation halo (\(\rmax=6\,\kpc\), \(\vmax=31\,\kms\)) with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\). Next we consider
the effect of softening as applied to our isolation halo.

@@

The two other key numerical parameters in \gadget{} are the tree-force
and time-step accuracy parameters. Specifically, \gadget{} opens a node
if \(M\,l/r^3 < \alpha |a|\), the node's mass is \(M\), distance from
the particle is \(r\), side-length is \(l\), and the particle's total
acceleration is \(a\). We adopt \(\alpha =0.005\). We also elect to use
adaptive time stepping with integration accuracy set to \(\eta=0.01\).
(particles must take time-steps smaller than
\(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)). The
lower panel of Fig.~\ref{fig:methods_convergence} tests changes to these
parameters. We include a model with smaller integration accuracy
(\(\eta=0.003\) in ``small timestep'') and a stricter tree-force
tolerance (\(\alpha=0.001\), ``high acc. force''). More precise
tolerances on these parameters does not affect the evolution within the
uncertainties of the final profile.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/orbit_converg_methods.png}}
\caption[Isolation method convergence]{A comparison of the final
profiles using different simulation methods for Sculptor's \smallperi{}
model. The benchmark model is our fiducial \(10^7\) particle run, and
all other models use \(10^5\) paritcles, with their ``converged radius''
marked by the black arrow and softening by the vertical bar(s).
\textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
lengths than the fiducial. \textbf{Bottom:} Models with more precise
timestep accuracy and gravitational force accuracy. \textbf{Summary:}
Except for the model with a larger softening length, all simulations
agree within uncertainties.}\label{fig:methods_convergence}
\end{figure}

>>>>

where \(h_{\rm grav}\) is the softening length, and \(N_{200}\) is the
number of particles within \(R_{200}\). This choice balances integration
time and only compromises resolution in the collisional regime. For our
isolation halo (\(\rmax=6\,\kpc\), \(\vmax=31\,\kms\)) with \(10^7\)
particles, this works out to be \(0.044\,{\rm kpc}\). Next, we consider
the effect of softening as applied to our isolation halo.

@@

The two other key numerical parameters in \gadget{} are the tree-force
and time-step accuracy parameters. Specifically, \gadget{} opens a node
if \(M\,l/r^3 < \alpha |a|\), where the node's mass is \(M\), the
distance from the particle is \(r\), the side-length is \(l\), and the
particle's total acceleration is \(a\). We adopt \(\alpha =0.005\). We
also elect to use adaptive time stepping with integration accuracy set
to \(\eta=0.01\). (particles must take time-steps smaller than
\(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for acceleration \(a\)). The
lower panel of Fig.~\ref{fig:methods_convergence} tests changes to these
parameters. We include a model with smaller integration accuracy
(\(\eta=0.003\) in ``small timestep'') and a stricter tree-force
tolerance (\(\alpha=0.001\), ``high acc. force''). More precise
tolerances on these parameters do not affect the evolution within the
uncertainties of the final profile.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/orbit_converg_methods.png}}
\caption[Isolation method convergence]{A comparison of the final
profiles using different simulation methods for Sculptor's \smallperi{}
model. The benchmark model is our fiducial \(10^7\) particle run, and
all other models use \(10^5\) particles, with their ``converged radius''
marked by the black arrow and softening by the vertical bar(s).
\textbf{Top:} Models with \(\sqrt{10}\) larger and smaller softening
lengths than the fiducial. \textbf{Bottom:} Models with more precise
timestep accuracy and gravitational force accuracy. \textbf{Summary:}
Except for the model with a larger softening length, all simulations
agree within uncertainties.}\label{fig:methods_convergence}
\end{figure}
